Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2013-04-03 HRUIonGPU/jrnl_article.tex
Final commit after submitting gpucapon paper to uffc.

\documentclass[journal]{IEEEtran}

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n]. \label{eq:z}
\end{align}
The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a better signal-to-noise ratio (SNR). Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

@@

\subsection{Beamspace Processing}\label{bs-capon}
Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{eq:butler} \\ 
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler or normalized DFT matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.

@@

In this section we give a brief discussion of GPU computing, laying the basis for our discussions on parallel implementation strategies. The Capon beamformers have been implemented using the CUDA GPU framework by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g. OpenCL. Independent of framework, the key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU. While the CPU has few cores, but logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to achieve maximum throughput.

@@

An SM has a limited amount of shared memory (or near-core cache) and registers. Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), a conservative use of registers and shared memory per thread is required. Each thread should perform enough instructions per transferred byte in order to hide memory latency, and the problem should be divided into enough threads in order to hide instruction latency. Global GPU memory transactions, in addition to CPU-GPU transfers, are slow and must be minimized. All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on both a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration. In the next section, we will describe how this mapping can be done for the ES-Capon beamformer.

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical.eps}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per thread-block if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring methods with reduced complexity.

@@

Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of the other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach. Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.

In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.

@@

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_bs.eps}}
\caption{Calculation of the sliding beamspace transformation. A group of $N_b$ threads are assigned a set of subarrays which they transform to beamspace in a sliding window fashion. In this example $N_b=3$ and $N_L\,\text{mod}\,N_b = 0$.}
\label{fig:gpulayoutbs}
\end{figure}

@@

In order to reduce the complexity of the matrix inversion problem the transformation must be performed before or after the covariance matrix has been estimated. If the transformation is applied before, one transformation is needed per subarray. This results in subarrays that no longer overlap. The amount of input data are therefore changed from $M$ to $N_LN_b$ elements per sample, but all remaining steps will get lower computational complexity as long as $N_b < L$.

@@

As mentioned, transforming subarrays into beamspace yields subarrays that no longer overlap. This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available. On the other hand, since $N_b$ should be small ($N_b < 32$), one thread assigned to each element in $\mat{\hat{R}}_\text{BS}$ is now an option. A kernel similar to the one described in \cite{Chen2011} has therefore been implemented. In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high GPU occupancy.

@@

Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}). In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations. The solution is to combine SBS with full computations. A number of thread groups is created that targets different sets of subarrays which they transform using the SBS algorithm. In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.

@@

To summarize, by transforming data to beamspace the complexity of all subsequent steps is reduced. Calculation of the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculation of weights and the beamformer output are both $O(N_b)$. The added complexity of SBS is $O(N_bL)$. To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.

>>>>

%\documentclass[journal]{IEEEtran}
\documentclass[11pt, draftcls, onecolumn]{IEEEtran}

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{eq:z}
\end{align}
where $(\cdot)^H$ is the Hermitian transpose. The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a better signal-to-noise ratio (SNR). Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

@@

\subsection{Beamspace Processing}\label{bs-capon}
Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{eq:butler} \\ 
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler or normalized discrete fourier transform (DFT) matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.

@@

In this section we give a brief discussion of GPU computing, laying the basis for our discussions on parallel implementation strategies. The Capon beamformers have been implemented using the CUDA GPU framework by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g. OpenCL. Independent of framework, the key point is that GPUs are designed to execute large sets of small concurrent tasks, like beamforming a pixel in a large image, much faster than a CPU. While the CPU has few cores, but logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and by that achieve maximum throughput.

@@

An SM has a limited amount of shared memory (or near-core cache) and registers. Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), a conservative use of registers and shared memory per thread is required. Each thread should perform enough instructions per transferred byte in order to hide memory latency, and the problem should be divided into enough threads in order to hide instruction latency. Global GPU memory transactions, in addition to CPU-GPU transfers, are slow and must be minimized. All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on both a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration. In the next two sections, we will describe how this mapping can be done for the Capon beamformers.

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical_2.eps}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per thread-block if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of the other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum supported subarray length is limited to 32 by this approach. Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.

In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.

@@

\begin{figure}
\centerline{\includegraphics[width=3.2in]{gfx/gpu_layout_bs_2.eps}}
\caption{Calculation of the sliding beamspace transform. A group of $N_b$ threads is assigned a set of subarrays which they transform to beamspace in a sliding window fashion. In this example $N_b=3$, $l$ is the current subarray and $k$ is the current beam.}
\label{fig:gpulayoutbs}
\end{figure}

@@

In order to reduce the complexity of the matrix inversion problem the transformation must be performed before or after the covariance matrix has been estimated. If the transformation is applied before, one transformation is needed per subarray. This results in subarrays that no longer overlap. The amount of input data are therefore changed from $M$ to $N_LN_b$ elements per sample, but all remaining steps should get lower computational complexity as long as $N_b < L$.

@@

As mentioned, transforming subarrays into beamspace yields subarrays that no longer overlap. This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available. On the other hand, since $N_b$ should be small ($N_b < 32$), one thread assigned to each element in $\mat{\hat{R}}_\text{BS}$ is now an option. A kernel similar to the one described in \cite{Chen2011} has therefore been implemented. In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high GPU occupancy. The computational complexity of this step is now $O(N_KN_LN_b^2)$, which for $L=M/2$ is less than the complexity of building $\mat{\hat{R}}$ in element space as long as $N_b < \sqrt{M}$.

@@

Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}). In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations. One solution, as depicted in Fig. \ref{fig:gpulayoutbs}, is to combine SBS with full computations. A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SBS algorithm. In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.

@@

To summarize, by transforming data to beamspace the complexity of all subsequent steps is potentially reduced. Calculation of the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculation of weights and the beamformer output are both $O(N_b)$. The added complexity of SBS is $O(N_bL)$. To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.
